---
title: "Computer Assignment 1"
author:
  - "Ville Sebastian Olsson (19911231-1999)"
  - "Arvind Guruprasad (19990212-6938)"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
---

```{r, echo=FALSE, results='hide', fig.show='hide'}
source("assignment1a.R")
list2env(ass1a, envir = .GlobalEnv)
knitr::opts_chunk$set(echo = FALSE, comment = "")
```

# Exercise 1:1

## 1. Percentages in favor and against legal abortion

We have a two-way contingency table with two categorical variables.
Let \(X\) denote the subject's gender, where \(X=1\) indicates a woman
and \(X=2\) indicates a man. Let \(Y\) denote the subject's opinion on
legal abortion, where \(Y=1\) indicates being in favor and \(Y=2\) indicates
being against.
Let \(n_{ij}\) denote the observed cell count in category \(X=i\), \(Y=j\).

The percentage is calculated as the number of subjects in a category
divided by the total number of subjects in the same row.

Percentage of women in favor:
\[\hat\pi_{1|1} = \frac{\hat\pi_{11}}{\hat\pi_{1+}}
= \frac{n_{11}/n}{n_{1+}/n}
= \frac{n_{11}}{n_{1+}} = \frac{309}{500}=0.618
\]
Percentage of women against:
\[\hat\pi_{2|1} = 1-\hat\pi_{1|1} = \frac{191}{500} = 0.382
\]
Percentage of men in favor:
\[\hat\pi_{1|2} = \frac{\hat\pi_{21}}{\hat\pi_{2+}}
= \frac{n_{12}/n}{n_{2+}/n}
= \frac{n_{12}}{n_{2+}} = \frac{319}{600}=0.532
\]
Percentage of men against:
\[\hat\pi_{2|2} = 1-\hat\pi_{1|2} = \frac{281}{600} = 0.468
\]

## 2. Different opinions on legal abortion?
Let:
\[\pi_{ij} = P(X=i,Y=j)\]
\[\pi_{i+}=P(X=i)\]
\[\pi_{+j}=P(Y=j)\]

### Hypotheses
We wish to investigate whether men and women have different opinions on legal
abortion.
This means investigating whether the events \(X=i\) and \(Y=j\) are independent
for all \(i,j\). As null hypothesis, we assume that they are independent:
\[H_0: \forall i,j\in\{1,2\}: \pi_{ij}=\pi_{i+}\pi_{+j}\]
Alternative hypothesis:
\[H_1: \exists i,j\in\{1,2\}: \pi_{ij} \neq \pi_{i+}\pi_{+j}\]

### Pearson's statistic
Let \(N_{ij}\) denote the random variable which \(n_{ij}\) is a realization of.

Pearson's \(\chi^2\) statistic is given by:
\[T_{\chi^2} = \sum_{i,j\in\{1,2\}} \frac{(N_{ij}-\hat\mu_{ij})^2}{\hat\mu_{ij}}\]
where \(\hat\mu_j\) is an estimate of the expected frequency for category \(j\), given by:
\[\hat\mu_{ij} = n\hat\pi_{ij} = n \hat\pi_{i+}\hat\pi_{+j} = n\frac{n_{i+}}{n}\frac{n_{+j}}{n}=\frac{n_{i+}n_{+j}}{n}\]
Estimates:
\[\hat\mu_{11} = \frac{500\cdot 628}{1100}=\frac{3140}{11}\]
\[\hat\mu_{12} = \frac{500\cdot 472}{1100}=\frac{2360}{11}\]
\[\hat\mu_{12} = \frac{600\cdot 628}{1100}=\frac{3768}{11}\]
\[\hat\mu_{12} = \frac{600\cdot 472}{1100}=\frac{2832}{11}\]
The statistic is, under \(H_0\), asymptotically \(\chi^2\) distributed with \(p-q = 3-2 = 1\)
degree of freedom, where \(p\) is the number of parameters in the unrestricted model
and \(q\) is the number of free parameters under \(H_0\).

Realization of Pearson's statistic:
\[t_{\chi^2} =
\frac{(309-\frac{3140}{11})^2}{3140/11}+
\frac{(191-\frac{2360}{11})^2}{2360/11}+
\frac{(319-\frac{3768}{11})^2}{3768/11}+
\frac{(281-\frac{2832}{11})^2}{2832/11}
\approx 8.30
\]

### LR statistic
The likelihood ratio (LR) statistic is given by:
\[T_{LR} = 2\sum_{i,j\in\{1,2\}} N_{ij}\ln\frac{N_{ij}}{\hat\mu_{ij}}\]
Like Person's statistic, the LR statistic is asymptotically \(\chi^2(1)\)
distributed under \(H_0\).

Realization of the LR statistic:
\[t_{LR} = 2\left(
309\cdot\ln\frac{309}{3140/11}+
191\cdot\ln\frac{191}{2360/11}+
319\cdot\ln\frac{319}{3768/11}+
281\cdot\ln\frac{281}{2832/11}
\right)
\approx 8.32\]

### Conclusion

The P-value for Pearson's \(\chi^2\) statistic is
approximately \(0.0040\) which is between significance levels \(0.001\) and
\(0.01\). Hence, we reject \(H_0\) in favor of \(H_1\) with high significance.

The P-value for the LR statistic is
approximately \(0.0039\) which is between significance levels \(0.001\) and
\(0.01\). Hence, we reject \(H_0\) in favor of \(H_1\) with high significance.

In conclusion, men and women differ in their opinions on legal abortion.

## 3. Odds ratio

### Odds
We are interested in comparing the odds of a woman being in favor of legal
abortion with the odds of a man being in favor of it.
Hence, we want to find the probabilities of \(Y\) conditioned on \(X\).

Odds for women:
\[\omega_1 = \frac{P(Y=1|X=1)}{P(Y=2|X=1)} = \frac{P(X=1,Y=1)/P(X=1)}{P(X=1,Y=2)/P(X=1)}=\frac{P(X=1,Y=1)}{P(X=1,Y=2)}
=\frac{\pi_{11}}{\pi_{12}}=\frac{N_{11}/n}{N_{12}/n}=\frac{N_{11}}{N_{12}}\]
Odds for men:
\[\omega_2 = \frac{P(Y=1|X=2)}{P(Y=2|X=2)} = \frac{P(X=2,Y=1)/P(X=2)}{P(X=2,Y=2)/P(X=2)}=\frac{P(X=2,Y=1)}{P(X=2,Y=2)}
=\frac{\pi_{21}}{\pi_{22}}=\frac{N_{21}/n}{N_{22}/n}=\frac{N_{21}}{N_{22}}\]
Odds ratio:
\[
\theta = \frac{\omega_1}{\omega_2}
= \frac{\pi_{11}/\pi_{12}}{\pi_{21}/\pi_{22}}
= \frac{\pi_{11}\pi_{22}}{\pi_{12}\pi_{21}}
\]
Estimated odds ratio:
\[
\hat\theta = \frac{\hat\omega_1}{\hat\omega_2}
= \frac{\hat\pi_{11}\hat\pi_{22}}{\hat\pi_{12}\hat\pi_{21}}
= \frac{(n_{11}/n)(n_{22}/n)}{(n_{12}/n)(n_{21}/n)}
= \frac{n_{11}n_{22}}{n_{12}n_{21}}
= \frac{309\cdot 281}{191 \cdot 319}
= \frac{86829}{60929}
\approx 1.43
\]

### Confidence interval

We start by constructing a confidence interval for the log-transformed odds
ratio:
\[\operatorname{CI}_{\ln\theta} = \left[\ln \hat\theta - z_{0.95} \operatorname{se}(\ln \hat\theta ),\ln \hat\theta  + z_{0.95} \operatorname{se}(\ln \hat\theta )\right]\]
with standard error:
\[\operatorname{se}(\ln\hat\theta)
= \sqrt{\widehat{\operatorname{Var}}(\ln\hat\theta)}
= \sqrt{\frac{1}{n_{11}}+\frac{1}{n_{12}}+\frac{1}{n_{21}}+\frac{1}{n_{22}}}
= \sqrt{\frac{1}{309}+\frac{1}{191}+\frac{1}{319}+\frac{1}{281}}
\approx 0.1231
\]
Realized confidence interval for \(\ln \theta\):
\[\operatorname{CI}_{\ln\theta}
\approx\left[\ln\frac{86829}{60929} - 1.96 \cdot 0.1231,\ln\frac{86829}{60929} + 1.96\cdot 0.1231\right]
\approx \left[0.1130,0.5955\right]
\]
We now exponentiate the limits to obtain the confidence interval for \(\theta\):
\[\operatorname{CI}_\theta \approx \left[e^{0.1130},e^{0.5955}\right] \approx [1.12,1.81]\]

### Interpretation
The estimated odds ratio between women and men is approximately 1.43, so
the odds are 43 \% higher for women to be in favor of legal abortion compared
to men.

The 95 % confidence interval for the odds ratio is approximately
\([1.12,1.81]\).
Since the interval does not contain 1, we conclude with 95 % confidence
that the odds of women being in favor of legal abortion are higher than for men.

## 4. Risk ratio

### Relative risk

The "risk" of women and men, respectively, being in favor of legal abortion:
\[r_1 = P(Y=1|X=1) = \frac{P(X=1,Y=1)}{P(X=1)} = \frac{\pi_{11}}{\pi_{1+}}\]
\[r_2 = P(Y=1|X=2) = \frac{P(X=2,Y=1)}{P(X=2)} = \frac{\pi_{21}}{\pi_{2+}}\]

Risk ratio:
\[r = \frac{r_1}{r_2} = \frac{\pi_{11}\pi_{2+}}{\pi_{21}\pi_{1+}}\]

Estimated relative risk:
\[\hat{r} = \frac{n_{11}/n_{1+}}{n_{21}/n_{2+}} = \frac{309/500}{319/600}
=\frac{1854}{1595} \approx 1.16\]

### Confidence interval
We start by constructing a confidence interval for the log-transformed odds
ratio:
\[\operatorname{CI}_{\ln r} = \left[
\ln \hat r - z_{0.95} \operatorname{se}(\ln \hat r ),
\ln \hat r + z_{0.95} \operatorname{se}(\ln \hat r )
\right]\]
with standard error:
\[\operatorname{se}(\ln\hat r)
= \sqrt{\widehat{\operatorname{Var}}(\ln\hat r)}
= \sqrt{\frac{1-n_{11}/n_{1+}}{n_{11}}+\frac{1-n_{21}/n_{2+}}{n_{21}}}
= \sqrt{\frac{1-309/500}{309}+\frac{1-319/600}{319}}
\]
Realized confidence interval for \(\ln r\):
\[\operatorname{CI}_{\ln r}
\approx [\ln\frac{1854}{1596}-1.96\cdot 0.05200,\frac{1854}{1596}+1.96\cdot 0.05200]
\approx [0.04793,0.2518]
\]
We now exponentiate the limits to obtain the confidence interval for \(r\):
\[\operatorname{CI}_r \approx [e^{0.04793},e^{0.2518}] \approx [1.05,1.29]\]

### Interpretation

The estimated risk ratio between women and men is approximately 1.16,
indicating that women are 16 % more likely than men to be in favor of legal
abortion.

The 95 % confidence interval for the risk ratio is approximately
\([1.05,1.29]\).
Since this interval does not contain 1, we conclude with 95 %
confidence that women are more likely to be in favor of legal abortion compared
to men.

## 5. Numerical result

Numerically computed percentages:
```{r}
percentages
```

Numerically computed realizations of the \(\chi^2\) statistic and LR statistic:
```{r}
square_statistics
```

Numerically computed odds ratio estimate and confidence interval:
```{r}
odds_ratio
```

Numerically computed risk ratio estimate and confidence interval:
```{r}
risk_ratio
```

These results are close approximations of the analytically calculated
values from earlier sections.

__

# Exercise 1:2

## 1. Analysis of admission data

We are given the admission data for the University of California, Berkeley grouped by gender. We are asked to create a table in R from the data provided and run the functions created earlier.

**Contingency table:**
```{r}
tab2
```

**Pearson's \(\chi^2\) statistic and LR statistic:**
```{r}
square_statistics2
```
Both tests indicate a very strong association between gender and admission.
The R function reports the p-values as 0.
In reality, the p-values are positive numbers very close to zero,
and R is unable to display them due to numerical precision limitations.
Since the p-values are lower than even the lowest common significance level of 0.001,
we reject the null hypothesis that gender and admission are independent. Thus, the data suggest that gender is significantly associated with admission.

**Odds Ratio:**

We seek the estimated odds of admission for men and the estimated odds of admission for women.
Dividing the former with the latter yields the estimated odds ratio for men compared to women:
\[\hat\theta = \frac{\hat\omega_1}{\hat\omega_2} = \frac{n_{11}/n_{12}}{n_{21}/n_{22}}\]
```{r}
odds_ratio2
```
The odds of admission for men are approximately **1.84** times higher than the odds of admission for women.
The confidence interval (1.62 to 2.08) does not include 1, indicating that this result is statistically significant.

**Risk Ratio:**

We seek the estimated probability of admission for men and the estimated probability of admission for women.
Dividing the former with the latter yields the estimated risk ratio for men compared to women:
\[\hat{r} = \frac{\hat{r}_2}{\hat{r}_1} = \frac{n_{11}/(n_{11}+n_{12})}{n_{21}/(n_{21}+n_{22})}\]

```{r}
risk_ratio2
```
The risk of admission for men is approximately **1.47** times higher than the risk of admission for women.
The confidence interval (1.35 to 1.59) does not include 1, indicating that this result is statistically significant.

**Summary**

Since the 95 \% confidence interval does not include 1,
this result is statistically significant.
The P-value is effectively zero for both the \(\chi^2\) and LR statistics, meaning that the result is very statistically significant. Men have both higher odds (1.84 times) and higher risk (1.47 times) of admission compared to women. This suggests a possible gender disparity in admission rates, favoring men. However, further investigation into other factors influencing admission (e.g., qualifications, department preferences, etc.) would be necessary to understand the full context.

## 2. Comparison with scaled data

Here, we would like to understand the effects of reducing the sample size of the data by dividing all the entries by a constant value. In this case, we are analyzing the statistics when we divide the input data by 10 and 100 respectively.

### Dividing the study population by 10

**Table:**
```{r}
tab2b
```
**Pearson's \(\chi^2\) statistic and LR statistic:**
```{r}
square_statistics2b
```

The test statistics (Pearson's \(\chi^2\) and LR statistic) have reduced by approximately a factor of 10, as expected because both are proportional to the size of the counts in the table. Despite the reduction, the association remains statistically significant (*p*<0.05), indicating that the relationship between gender and admission is strong enough to persist even with smaller sample sizes.

**Odds ratio:**
```{r}
odds_ratio2b
```

The odds ratio remains nearly identical (~1.84) as it is a relative measure and is unaffected by the scale of the data. However, the confidence interval widened slightly, reflecting uncertainty due to the smaller sample size.  

**Risk ratio:**
```{r}
risk_ratio2b
```

Similar to the odds ratio, the risk ratio (~1.46) is identical given the relative nature of the statistic. And due to the smaller sample size, the confidence interval widens more noticeably, indicating reduced precision. 

### Dividing the study population by 100

**Table:**
```{r}
tab2c
```
**Pearson's \(\chi^2\) statistic and LR statistic:**
```{r}
square_statistics2c
```

The test statistics have significantly decreased, as expected, with smaller sample sizes.
The p-value is no longer significant (*p*>0.05), meaning the test fails to detect a significant association between gender and admission. This reflects the reduced power of the tests when the sample size is too small.

**Odds ratio:**
```{r}
odds_ratio2c
```

The odds ratio remains consistent with the original result, but the confidence interval is much wider, spanning from less than 1 to well over 5. This indicates high uncertainty due to the small sample size.

**Risk ratio:**
```{r}
risk_ratio2c
```

Like the odds ratio, the risk ratio is consistent but with a much wider confidence interval, again reflecting the increased uncertainty.

Dividing the sample size by 100 further demonstrates the limitations of small data sets in hypothesis testing and estimation. While odds and risk ratios remain unchanged in magnitude, the increased uncertainty renders these measures less informative.
This highlights the importance of sufficient sample sizes for reliable statistical inference. With small data sets, significant effects might go undetected, and confidence in parameter estimates is greatly diminished.

## 3. Custom two-way table

Our objective is to find a two-way contingency table of the form:

\[
\begin{array}{|c|c|c|c|}
\hline
 & Y=1 & Y=2 \\
\hline
X=1 & n_{11} & n_{12} \\
\hline
X=2 & n_{21} & n_{22} \\
\hline
\end{array}
\]

under the constraints:
\[\hat\theta = \frac{n_{11}/n_{12}}{n_{21}/n_{22}} \in (0.99,1.01)\]
\[\text{p-value of }t_{\chi^2} < 0.05\]
\[\text{p-value of }t_{LR} < 0.05\]
where \(\hat\theta\) is the estimated odds ratio, \(t_{\chi^2}\) is the realized Pearson's \(\chi^2\) statistic, and \(t_{LR}\) is the realized LR statistic.

Based on the provided constraints where the sample odds ratio is between 0.99 and 1.01 and the corresponding odds ratio should differ from 1, we can choose the following values:
\[n_{11} = 500000\]
\[n_{12} = 501500\]
\[n_{21} = 500500\]
\[n_{22} = 499000\]

**Table:**
```{r}
tab2d
```
**Odds ratio:**
```{r}
odds_ratio2d
```
The estimated odds ratio is approximately 99.4 \% which is within the bounds of \((0.99,1.01)\).

**Pearson's \(\chi^2\) statistic and LR statistic:**
```{r}
square_statistics2d
```
The P-value for both statistics is approximately 3.3 \%. The p-value is less than 0.05, so the result meets the conventional threshold for statistical significance.

Given that the estimated odds ratio is so close to 1, the result may not have
any practical significance in most real-world contexts even though the result is
statistically significant.
For example, if the odds ratio represents the admission rate for men and women,
the difference is so small that it is not a noteworthy result.
In situations with very large sample sizes, even trivial differences can achieve statistical significance due to the high power of the tests.

The results in such instances would involve providing the entire context, discussing both statistical and practical significance. To report this result, one could utilize visualizations to indicate its lack of practical significance. This result could be expressed using for example a forest plot, a bar plot of the relative frequencies, or a confidence interval for the odds ratio.
Hence, statistical significance alone does not imply real-world importance.