---
title: "Computer Assignment 3"
author:
  - "Ville Sebastian Olsson (19911231-1999)"
  - "Arvind Guruprasad (19990212-6938)"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
---

```{r, echo=FALSE, results='hide', fig.show='hide'}
source("assignment3.R")
list2env(ass, envir = .GlobalEnv)
knitr::opts_chunk$set(echo = FALSE, comment = "")
```

# Exercise 3:1

## 1. Model selection

The data comes from a study reported by Wermuth (1976), which was collected in a birth clinic. The dataset includes the following variables:

- **Mother's age (X):** `<30` and `30+`.
- **Smoking habits (Y):** `<5 cigarettes/day` and `5+ cigarettes/day`.
- **Gestational age (Z):** `<260 days` and `≥260 days`.
- **Child survival (V):** Binary outcome indicating survival (Yes/No).

The goal is to build log-linear models to understand the associations between these categorical variables.

To generate a table, we follow this process:

1. Read the CSV data
2. Fit the saturated model
3. For each combination of interaction terms, fit a model
4. For each sub-model, calculate deviance, AIC and p-value with respect to the saturated model

Result:
```{r}
model_table
```


## 2.

The table contains the following information which helps us
select the best model:

**1. Deviance**

- Measures the difference in goodness-of-fit between the submodel and the saturated model.
- A low deviance indicates a good fit.

**2. Degrees of Freedom**

- Reflects the number of parameters removed from the saturated model.
- Higher df means a simpler model.

**3. p-value**

- Tests whether the submodel is significantly worse than the saturated model.
- If \( p > 0.05 \), the simpler model is adequate.

**4. AIC**

- Evaluates the trade-off between model complexity and fit.
- Select the model with the lowest AIC.

```{r}
model_table[order(model_table$AIC),]
```

The table above has been sorted by AIC in ascending order,
to gauge which of the simpler model would be effective with
respect to the saturated model.

We should select the model that satisfies two criteria:

* Minimal AIC value
* The p-value is greater than 0.05

Based on the table above, we want to select model 6,
which includes the interaction terms `XZ, XV, ZV, Y`.

# 3.

The model we selected looks like this:

```{r}
best_model
```

Given the lack of three-way interactions, this model
is much simpler relative to the relative model and the
other three-way interaction model while retaining
its ability to fit the data reasonably well.

```{r}
associations
```


1. **Main Effects**:
   - \(x\), \(z\), and \(y\) have significant **negative associations** with the outcome.
   - \(v\) has a significant **positive association** with the outcome.

2. **Interaction Effects**:
   - \(x : z\) and \(x : y\): Significant **negative interactions**, indicating their combined effects reduce the odds of the outcome beyond individual effects.
   - \(z : v\): A significant **positive interaction**, with their combined effects dramatically increasing the odds of the outcome.

Detailed Interpretation:

1. **Main Effects**:
   - \(x\): A significant negative association (\(OR = 0.51\), \(CI: 0.43 \, - \, 0.60\)), meaning a 1-unit increase in \(x\) reduces the odds by 49%.
   - \(z\): Similarly, \(z\) has a significant negative association (\(OR = 0.47\), \(CI: 0.33 \, - \, 0.66\)), reducing the odds by 53%.
   - \(y\): A strong negative association (\(OR = 0.12\), \(CI: 0.11 \, - \, 0.13\)), where a 1-unit increase reduces the odds by 88%.
   - \(v\): A significant positive association (\(OR = 4.93\), \(CI: 4.00 \, - \, 6.09\)), meaning \(v\) increases the odds nearly fivefold.

2. **Interaction Effects**:
   - \(x : z\): A significant negative interaction (\(OR = 0.79\), \(CI: 0.66 \, - \, 0.94\)), indicating the combined effect of \(x\) and \(z\) decreases odds by 21% beyond their individual effects.
   - \(x : y\): Another significant negative interaction (\(OR = 0.67\), \(CI: 0.55 \, - \, 0.81\)), reducing odds by 33% for combined increases in \(x\) and \(y\).
   - \(z : v\): A highly significant positive interaction (\(OR = 27.88\), \(CI: 19.43 \, - \, 40.01\)), suggesting that the joint presence of \(z\) and \(v\) increases the odds by nearly 28 times.

# 4.
